---
title: "alpha_analysis"
format: html
---

## Alpha Analysis

**Obj 3**: What is the variation of alpha values across temperature and lifestage

We expect that alpha will increase with temperature and temparature will have a greater impact at the transitional saltwater lifestages (saltwater smolt and subadult).

-   First we have to do some data tidying after we load in the data.

```{r}
#load in libraries
library(tidyverse)

#read in relevant datafiles
alpha_int <- read.csv("alpha_int.csv")

#### the following code is data tidying and may be duplicative ####



#add column "Salinity" identifying if a fish was saltwater or freshwater
alpha_int <- alpha_int %>%
  mutate(salinity = case_when(
    sal_ppt == 35 ~ "Saltwater",
    sal_ppt == 0  ~ "Freshwater"))

# add column classifying if fish was ran in fresh or saltwater and what lifestage it is
# use lubridate on datetime column
alpha_int$datetime <- ymd_hms(alpha_int$datetime)

# use dates of exp to determine if saltwater juvenile, freshwater juvenile or subadult
sj_cutoff <- ymd_hms("2025-07-24 00:00:00", tz = "UTC")
fj_cutoff <- ymd_hms("2025-08-03 00:00:00", tz = "UTC")
alpha_int <- alpha_int %>%
  mutate(lifestage = case_when(
    datetime < sj_cutoff ~ "saltwater smolt",
    datetime < fj_cutoff ~ "freshwater smolt",  # anything >= saltwater smolt cutoff but < freshwater cutoff
    TRUE ~ "subadult"                   # everything else
  )) 
# change lifestage column to a factor
alpha_int$lifestage <- as.factor(alpha_int$lifestage)

# convert datetime to just date
 alpha_int <- alpha_int %>% 
    mutate(date = as.Date(datetime)) %>% 
    select(-datetime)


# average the alpha and temp at for each fish
alpha_values<- alpha_int %>% 
  group_by(fish_id, salinity,lifestage,date) %>% 
  summarise(avg_alpha= mean(alpha_int_mgo2_kg_h_kPa),
            avg_temp = mean(temp_c)) %>% 
  ungroup()
```

-   Here we read in the post mortem data and then combine it with our alpha values df.

```{r}
post_mort <- read.csv("meta_postmort.csv")

# add relevant columns to the combined df
alpha_values <- alpha_values %>% 
  left_join(post_mort %>% 
              select(fish_id, cf, hsi, l_perc_dw, t_perc_dw, posttrial_wet_weight_g),
            by = "fish_id")
```

## Plotting

Here we will plot potentially explanatory variables for alpha values, most interesting is the life stage and temperature. We will also look at the post mortem data to see if they have any correlation.

Plots include the following variables: temperature, lifestage, condition factor, liver % dry weight, and tissue % dry weight.

```{r}
#plot alpha vs temp by lifestage
ggplot(alpha_values, aes(x = avg_temp, y = avg_alpha, color = lifestage)) +
  geom_point(alpha = 0.7)+
  labs(
    title = "Alpha vs Temperature by Lifestage",
    x = "Temperature (°C)",
    y = "Alpha (mgO2kg-1h-1kPa-1)") +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Set2", name = "Lifestage", labels = c("Freshwater smolt", "Saltwater smolt", "Subadult"))
  
#plot alpha vs condition factor by lifestage
ggplot(alpha_values, aes(x = cf, y = avg_alpha, color = lifestage)) +
  geom_point(alpha = 0.7)+
  labs(
    title = "Alpha vs CF by Lifestage",
    x = "Condition Factor",
    y = "Alpha (mgO2kg-1h-1kPa-1)") +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Set2", name = "Lifestage", labels = c("Freshwater smolt", "Saltwater smolt", "Subadult"))

#plot alpha vs % liver dry weight 
# higher % liver dry weight indicates decreased stress
ggplot(alpha_values, aes(x = l_perc_dw, y = avg_alpha, color = lifestage)) +
  geom_point(alpha = 0.7)+
  labs(
    title = "Alpha vs Liver % Dryweight by Lifestage",
    x = "Liver % dry weight",
    y = "Alpha (mgO2kg-1h-1kPa-1)") +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Set2", name = "Lifestage", labels = c("Freshwater smolt", "Saltwater smolt", "Subadult"))


# alpha vs % tissue dry weight
ggplot(alpha_values, aes(x = t_perc_dw, y = avg_alpha, color = lifestage)) +
  geom_point(alpha = 0.7)+
  labs(
    title = "Alpha vs % Tissue dry weight by Lifestage",
    x = "% tissue dry weight",
    y = "Alpha (mgO2kg-1h-1kPa-1)") +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Set2", name = "Lifestage", labels = c("Freshwater smolt", "Saltwater smolt", "Subadult"))


# plot alpha vs hsi, note there are no subadult's because they did not get an hsi value

ggplot(alpha_values, aes(x = hsi, y = avg_alpha, color = lifestage)) +
  geom_point(alpha = 0.7)+
  labs(
    title = "Alpha vs hsi by Lifestage",
    x = "HSI",
    y = "Alpha (mgO2kg-1h-1kPa-1)") +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Set2", name = "Lifestage", labels = c("Freshwater smolt", "Saltwater smolt", "Subadult"))
```

**CF**: It looks like there could be some correlation between alpha and condition factor, but interestingly it is the opposite trend of what you would expect. Condition factor (CF)is a measure of a fish's mass/length, so a higher CF means that the fish is healthier. We would expect that a higher CF would mean a higher alpha because the fish is supposedly more physiologically fit.

**HSI**: It looks like this has an interesting trend because the HSI is lower for the saltwater smolts which could indicate that they are more physiologically stressed. However, again it has an unexpected inverse relationship with alpha where would expect it to be a positive relationship. HSI could be impacted by the saltwater transition which could be introducing additional influence.

For the scope of this project we will stick to temperature and lifestage.

## Checking if Data are Normally Distributed

-   Shapiro test will tell us if the data is normally distributed. Null hypothesis is that data is normally distributed. Would be confirmed by a p \< 0.05.

-   qqPlot will visualize if the data falls within a normal distribution. We want data points within the blue bounds.

```{r}
#test for normalcy
shapiro.test(alpha_values$avg_alpha)
qqPlot(alpha_values$avg_alpha)

```

The Shapiro test shows that we do not have normally distributed data, but the qq-plot shows that there is only slight deviation from a normal distribution with no major outliers. We will assume a normal distribution.

## Modeling

We will have four models combining lifestage and temperature across additive and interactive effects. Then we will compare them to each other with AIC and BIC to determine the best model.

```{r}
lm1 <- lm(avg_alpha ~ lifestage, alpha_values)
lm2 <- lm(avg_alpha ~ avg_temp, alpha_values)
lm3 <- lm(avg_alpha ~ lifestage + avg_temp, alpha_values)
lm4 <- lm(avg_alpha ~ lifestage + avg_temp + lifestage:avg_temp, alpha_values)

#AIC/BIC evaluation
#df that shows AIC and BIC results
evaluation <- data.frame(AIC(lm1, lm2, lm3, lm4)) %>%
  mutate(BIC = BIC(lm1, lm2, lm3, lm4)$BIC) %>%
  arrange(BIC)
```

-   From the AIC and BIC results we can see that lm4 is the best fit model. This contains lifestage, temperature and the interactive effect of lifestage and temperature.

-   Now lets look at summary statistics about this model

```{r}
# we can visualize to make sure that data looks evenly distributed
# anova allows us to determine if the variables have a significant effect
plot(lm4)
Anova(lm4)
```

-   We can see a good distribution across our four plots and the Anova shows that we have significance of our variables.

-   Let's look at the summary to understand the significance.

```{r}
summary(lm4)
```

**Intercept**: Both the saltwater smolts and subadult fish have significantly lower intercepts of alpha compared to freshwater smolts.

**Temperature**: Alpha increases by 0.57 for every 1 degree Celsius increase. This supports our initial hypothesis from earlier.

**Temperature conditional on lifestage:** Finally, we will look at how the interactive effect of temperature and lifestage impacts alpha. From the model we see that for the saltwater smolts, alpha increases by 1.22 more than freshwater smolts for every 1 degree Celsius increase. For the subadults their alpha increases by 0.496 more than freshwater smolts for every 1 degree Celsius increase. This supports our initial hypothesis too.

Interestingly, the saltwater smolt lifestage is impacted more by temperature than the subadult lifestage. This could be because the smolt lifestage is a very stressful transitional period as it is acclimating to saltwater.

## Visualize the Model

-   The best way to understand this model is to visualize it

```{r}
#visualize the model
library(visreg)
visreg(lm4, "avg_temp", by = "lifestage",)
```

We can see that the slope of saltwater smolts is the steepest meaning that temperature has a greater impact on it than the other lifestages. Subadult fish also have an increased slope compared to freshwater smolts, although not as steep as saltwater smolts so it is also impacted by temperature more than the freshwater smolts.

## Predicting Data

-   We can look at predicted data to help us understand our model

-   First we create new data based based on the range of data we used to create our predicted model. Then we will use the model to predict using that data.

-   Note that for these plots the line represents the predicted trend, and the points represent the actual data, this allows us to see how well our model aligns with the actual data.

```{r}
# creating new data of all combinations of lifestage and temp
newdata <- expand.grid(lifestage = unique(alpha_values$lifestage),
                       avg_temp = seq(min(alpha_values$avg_temp,na.rm=T),
                                      max(alpha_values$avg_temp,na.rm=T),
                                      length.out=100))


# how does model4 predict this new data
prdata <- predict(lm4,newdata = newdata, se.fit = T)

#creating plot of the new data fit
newdata$fit <- prdata$fit
newdata$lcl <- prdata$fit - prdata$se.fit*2 #lower cl, -fit x2
newdata$ucl <- prdata$fit + prdata$se.fit*2 #upper cl, +fit x2

#plot of this fit with the ucl and lcl
newdata %>%
  ggplot(aes(avg_temp, fit, fill=lifestage)) +
    geom_point(data = alpha_values, aes(avg_temp,avg_alpha ,col = lifestage)) +
    geom_line(aes(col = lifestage)) +
    geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.3) +
    labs(y = "Alpha (mgO2kg-1h-1kPa-1)", x = "Temperature (°C)")

# also make this same plot but facet wrapped so you can see the individual lifestages
newdata %>%
  ggplot(aes(avg_temp,fit, fill=lifestage)) +
    geom_point(data = alpha_values, aes(avg_temp,avg_alpha ,col = lifestage)) +
    geom_line(aes(col = lifestage)) +
    geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.3) +
    facet_wrap(~lifestage) +
    labs(y = "Alpha (mgO2kg-1h-1kPa-1)", x = "Temperature (°C)") +
    theme(legend.position = "none")
```

## Visualizing Variation

-   Lets create a plot that helps visualize the range of alpha across temperatures between the different lifestages.

-   Note that this is done with the predicted data, not the actual data

```{r}
newdata %>% 
  ggplot(aes(lifestage, fit, colour = avg_temp)) +
    geom_point() +
    scale_color_gradient(name = "Temperature", low = "blue", high = "red") +
      labs(title = "Predicted alpha", x = "Life stage", y = "Estimated alpha") +
      theme(plot.title = element_text(hjust = 0.5))
```

-   This is what it would look like with the actual data

```{r}
alpha_values %>% 
  ggplot(aes(lifestage, avg_alpha, colour = avg_temp)) +
    geom_point() +
    scale_color_gradient(name = "Temperature", low = "blue", high = "red") +
      labs(title = "Alpha range", x = "Life stage", y = "alpha") +
      theme(plot.title = element_text(hjust = 0.5))
```

We see that for the saltwater smolt lifestage, temperature has causes a greater variation in alpha ranges when compared to the freshwater smolt lifestage. This is present in the subadult lifestage as well, but not as strong of an effect. This illustrates the impact of temperature on the transitional lifestages as the enter the ocean.
