---
title: "respirometry_data_analysis"
format: html
editor: visual
---

# Install and Load Packages

```{r}
#install.packages(respirometry)
#install.packages(car)
#install.packages(visreg)

library(tidyverse)
library(respirometry)
library(stats)
library(car)
library(lubridate)
```

# Data Tidying

## Tidying the Alpha Intermittent data frame

-   This section will calculate the avg of the 3 highest alpha values from the intermittent experiment for each fish

```{r}
alpha_int <- read.csv("alpha_int.csv")

#add column "Salinity" identifying if a fish was saltwater or freshwater
alpha_int <- alpha_int %>%
  mutate(salinity = case_when(
    sal_ppt == 35 ~ "Saltwater",
    sal_ppt == 0  ~ "Freshwater"))

# add column classifying if fish was ran in fresh or saltwater and what lifestage it is
# use lubridate on datetime column
alpha_int$datetime <- ymd_hms(alpha_int$datetime)

# use dates of exp to determine if saltwater juvenile"SJ", freshwater juvenile"FJ" or subadult saltwater"SS"
sj_cutoff <- ymd_hms("2025-07-24 00:00:00", tz = "UTC")
fj_cutoff <- ymd_hms("2025-08-03 00:00:00", tz = "UTC")
alpha_int <- alpha_int %>%
  mutate(lifestage = case_when(
    datetime < sj_cutoff ~ "SJ",
    datetime < fj_cutoff ~ "FJ",  # anything >= SJ cutoff but < FJ cutoff
    TRUE ~ "SS"                   # everything else
  )) 
# change lifestage column to a factor
alpha_int$lifestage <- as.factor(alpha_int$lifestage)

#keep only the highest alpha value
#this is to test if there is any difference between the single highest and the avg of the highest 3 values
highest_alpha_int <- alpha_int %>%
  group_by(fish_id) %>%
  slice_max(order_by = alpha_int_mgo2_kg_h_kPa, n = 1) %>%
  ungroup()


#avg the alpha_max value
#create new df with the avg and the highest alpha value

alpha_values_int<- alpha_int %>% 
  group_by(fish_id, salinity) %>% 
  summarise(avg_alpha_int= mean(alpha_int_mgo2_kg_h_kPa),
            avg_temp_int = mean(temp_c)) %>% 
  left_join(highest_alpha_int %>%
              select(fish_id, alpha_int_mgo2_kg_h_kPa,lifestage), by = "fish_id") %>% 
  rename(highest_alpha_int = alpha_int_mgo2_kg_h_kPa) %>% 
  ungroup()

```

## Tidying the Alpha Closed data frame

-   This section will calculate the avg of the 3 highest alpha values from the closed experiment for each fish

-   \*Note: Not all fish have a closed experiment because some died during the intermittent experiment

```{r}
# read in closed alpha values
alpha_closed <- read.csv("alpha_close.csv")

#add in salinity
alpha_closed <- alpha_closed %>%
  mutate(salinity = case_when(
    sal_ppt == 35 ~ "Saltwater",
    sal_ppt == 0  ~ "Freshwater"))

#keep highest alpha value
#this is to test if there is any difference between the single highest and the avg of the highest 3 values
highest_alpha_closed <- alpha_closed %>%
  group_by(fish_id) %>%
  slice_max(order_by = alpha_close_mgo2_kg_h_kPa, n = 1) %>%
  ungroup()

#avg alpha value and make new df with both avg and highest values
alpha_values_closed<- alpha_closed %>% 
  group_by(fish_id, salinity) %>% 
  summarise(avg_alpha_closed= mean(alpha_close_mgo2_kg_h_kPa),
            avg_temp_closed = mean(temp_c)) %>% 
  left_join(highest_alpha_closed %>%
              select(fish_id, alpha_close_mgo2_kg_h_kPa), by = "fish_id") %>% 
  rename(highest_alpha_closed = alpha_close_mgo2_kg_h_kPa) %>% 
  ungroup()

```

## Combining Alpha Values

-   This section will combine alpha values from the intermittent and closed portions of the experiment

-   \*Note: not all fish will have both values because some died during the intermittent experiment

    ```{r}
    #creating new data frame, joining both alpha values
    alpha_values_combined <- alpha_values_int %>% 
      left_join(alpha_values_closed %>% 
                  select(fish_id, avg_alpha_closed, avg_temp_closed),
                by = "fish_id")

    # calculate difference between alpha_closed and alpha_int
    # avg temp values of closed and int values, can do a simple avg because there are equal # of values in closed and int
    alpha_values_combined <- alpha_values_combined %>% 
      mutate(alpha_diff_avg = avg_alpha_int -avg_alpha_closed,
             temp_avg = ((avg_temp_closed + avg_temp_int)/2)) %>% 
      select(-c(avg_temp_int, avg_temp_closed))
    ```

## Tidying the Post-Mortem data frame 

-   Here we are looking at other variables and interactions

-   Lets start by looking at the post mortem data

    -   condition factor (CF)

    -    hepatosomatic index (HSI)

    -   \% tissue and liver dry weight

    ```{r}
    # Reading and naming data frame
    post_mort <- read.csv("meta_postmort.csv")

    # add relevant columns to the combined df
    alpha_values_combined <- alpha_values_combined %>% 
      left_join(post_mort %>% 
                  select(fish_id, cf, hsi, l_perc_dw, t_perc_dw, posttrial_wet_weight_g, fish_mort_pcrit),
                by = "fish_id")

    # removing fish that died during pcrit trial
    # a total of 10 fish removed, 2 were already removed due to acclimation flag
    alpha_values_combined <- alpha_values_combined %>% 
      filter(fish_mort_pcrit != "Y")


    #adding date to do some exploratory plotting
    alpha_values_combined <- alpha_values_combined %>% 
      left_join(highest_alpha_int %>% 
                  select(fish_id, datetime),
                by = "fish_id") %>% 
      mutate(date = as.Date(datetime,)) %>% 
      select(-datetime)
    ```

# Exploratory Data Analysis

## T-Test

-   Here we are going to run a t-test with avg alpha_int and avg_alpha_closed

-   This will tell us if the two values are different, essentially calculating the difference between them

```{r}
# running t-test with avg alpha_int and avg_alpha_closed
ttest <- alpha_values_combined %>% 
  with(t.test(avg_alpha_int, avg_alpha_closed, paired = TRUE, na.action = na.omit))
print(ttest)

# avg_alpha_int is statistically different than avg_alpha_closed
# mean difference 4.48
# therefore we should use the alpha values from the intermittent experiment
```

## Checking if Data are Normally Distributed

```{r}
shapiro.test(alpha_values_combined$alpha_diff_avg)
qqPlot(alpha_values_combined$alpha_diff_avg)
```

## Producing Plot of Pairwise Correlations Among Predictors

```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y,use = "all.obs",method = "spearman"))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt) }


temp <- alpha_values_combined %>%
  filter(!is.na(alpha_diff_avg)) %>%
  ungroup() %>%
  mutate(ww_log = log(posttrial_wet_weight_g)) %>%
  select(ww_log,temp_avg)

# plot correlations
pairs(temp,lower.panel = panel.smooth, upper.panel = panel.cor,
      gap=0, row1attop=FALSE)
```

# Explanatory Data Analysis
