---
title: "respirometry_data_analysis"
format: html
editor: visual
---

# Install and Load Packages

```{r}
#install.packages(respirometry)
#install.packages(car)
#install.packages(visreg)

library(tidyverse)
library(respirometry)
library(stats)
library(car)
library(lubridate)
```

# Data Tidying

## Tidying the Alpha Intermittent data frame

-   This section will calculate the avg of the 3 highest alpha values from the intermittent experiment for each fish

```{r}
alpha_int <- read.csv("alpha_int.csv")

#add column "Salinity" identifying if a fish was saltwater or freshwater
alpha_int <- alpha_int %>%
  mutate(salinity = case_when(
    sal_ppt == 35 ~ "Saltwater",
    sal_ppt == 0  ~ "Freshwater"))

# add column classifying if fish was ran in fresh or saltwater and what lifestage it is
# use lubridate on datetime column
alpha_int$datetime <- ymd_hms(alpha_int$datetime)

# use dates of exp to determine if saltwater juvenile "SJ", freshwater juvenile "FJ" or subadult saltwater "SS"
sj_cutoff <- ymd_hms("2025-07-24 00:00:00", tz = "UTC")
fj_cutoff <- ymd_hms("2025-08-03 00:00:00", tz = "UTC")
alpha_int <- alpha_int %>%
  mutate(lifestage = case_when(
    datetime < sj_cutoff ~ "saltwater smolt",
    datetime < fj_cutoff ~ "freshwater smolt",  # anything >= SJ cutoff but < FJ cutoff
    TRUE ~ "subadult"                   # everything else
  )) 
# change lifestage column to a factor
alpha_int$lifestage <- as.factor(alpha_int$lifestage)

#keep only the highest alpha value
#this is to test if there is any difference between the single highest and the avg of the highest 3 values
highest_alpha_int <- alpha_int %>%
  group_by(fish_id) %>%
  slice_max(order_by = alpha_int_mgo2_kg_h_kPa, n = 1) %>%
  ungroup()


#avg the alpha_max value
#create new df with the avg and the highest alpha value

alpha_values_int<- alpha_int %>% 
  group_by(fish_id, salinity) %>% 
  summarise(avg_alpha_int= mean(alpha_int_mgo2_kg_h_kPa),
            avg_temp_int = mean(temp_c)) %>% 
  left_join(highest_alpha_int %>%
            select(fish_id, alpha_int_mgo2_kg_h_kPa,lifestage),
            by = "fish_id") %>% 
  rename(highest_alpha_int = alpha_int_mgo2_kg_h_kPa) %>% 
  ungroup()

```

## Tidying the Alpha Closed data frame

-   This section will calculate the avg of the 3 highest alpha values from the closed experiment for each fish

-   \*Note: Not all fish have a closed experiment because some died during the intermittent experiment

```{r}
# read in closed alpha values
alpha_closed <- read.csv("alpha_close.csv")

#add in salinity
alpha_closed <- alpha_closed %>%
  mutate(salinity = case_when(
          sal_ppt == 35 ~ "Saltwater",
          sal_ppt == 0  ~ "Freshwater"))

#keep highest alpha value
#this is to test if there is any difference between the single highest and the avg of the highest 3 values
highest_alpha_closed <- alpha_closed %>%
  group_by(fish_id) %>%
  slice_max(order_by = alpha_close_mgo2_kg_h_kPa, n = 1) %>%
  ungroup()

#avg alpha value and make new df with both avg and highest values
alpha_values_closed<- alpha_closed %>% 
  group_by(fish_id, salinity) %>% 
  summarise(avg_alpha_closed = mean(alpha_close_mgo2_kg_h_kPa),
            avg_temp_closed = mean(temp_c)) %>% 
  left_join(highest_alpha_closed %>%
              select(fish_id, alpha_close_mgo2_kg_h_kPa),
                     by = "fish_id") %>% 
  rename(highest_alpha_closed = alpha_close_mgo2_kg_h_kPa) %>% 
  ungroup()

```

## Combining Alpha Values

-   This section will combine alpha values from the intermittent and closed portions of the experiment

-   \*Note: not all fish will have both values because some died during the intermittent experiment

    ```{r}
    #creating new data frame, joining both alpha values
    alpha_values_combined <- alpha_values_int %>% 
      left_join(alpha_values_closed %>% 
                select(fish_id, avg_alpha_closed, avg_temp_closed),
                by = "fish_id")

    # calculate difference between alpha_closed and alpha_int
    # avg temp values of closed and int values, can do a simple avg because there are equal # of values in closed and int

    alpha_values_combined <- alpha_values_combined %>% 
      mutate(alpha_diff_avg = avg_alpha_int -avg_alpha_closed,
             temp_avg = ((avg_temp_closed + avg_temp_int)/2)) %>% 
      select(-c(avg_temp_int, avg_temp_closed))
    ```

## Tidying the Post-Mortem data frame

-   Here we are looking at other variables and interactions

-   Lets start by looking at the post mortem data

    -   condition factor (CF)

    -   hepatosomatic index (HSI)

        -   Sub adult fish do not have an HSI value

    -   \% tissue and liver dry weight

    ```{r}
    # Reading and naming data frame
    post_mort <- read.csv("meta_postmort.csv")

    # add relevant columns to the combined df
    alpha_values_combined <- alpha_values_combined %>% 
      left_join(post_mort %>% 
                  select(fish_id, cf, hsi, l_perc_dw, t_perc_dw, posttrial_wet_weight_g, fish_mort_pcrit),
                by = "fish_id")

    # removing fish that died during pcrit trial
    # a total of 10 fish removed, 2 were already removed due to acclimation flag
    alpha_values_combined <- alpha_values_combined %>% 
      filter(fish_mort_pcrit != "Y")


    # adding date to do some exploratory plotting
    alpha_values_combined <- alpha_values_combined %>% 
      left_join(highest_alpha_int %>% 
                select(fish_id, datetime),
                by = "fish_id") %>% 
      mutate(date = as.Date(datetime)) %>% 
      select(-datetime)
    ```

# Exploratory Data Analysis

## T-Test

-   Here we are going to run a t-test with avg alpha_int and avg_alpha_closed

-   This will tell us if the two values are different, essentially calculating the difference between them

```{r}
# running t-test with avg alpha_int and avg_alpha_closed

ttest <- alpha_values_combined %>% 
  with(t.test(avg_alpha_int, avg_alpha_closed, paired = TRUE, na.action = na.omit))
print(ttest)

# avg_alpha_int is statistically different than avg_alpha_closed
# mean difference: 4.48
# therefore we will use the alpha values from the intermittent experiment
```

## Checking if Data are Normally Distributed

-   Shapiro test will tell us if the data is normally distributed. Null hypothesis that is that data is normally distributed. Would be confirmed by a p \< 0.05.

-   qqPlot will visualize if the data falls within a normal distribution. We want data points within the blue bounds.

```{r}
shapiro.test(alpha_values_combined$alpha_diff_avg)
qqPlot(alpha_values_combined$alpha_diff_avg)
```

From the Shapiro test, we do not have normally distributed data. However, this is only due to two outlying data points so we will go ahead and assume a normal distribution.

## Producing Plot of Pairwise Correlations Among Predictors

```{r}
# Need to ask Ben how this works

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y,use = "all.obs",method = "spearman"))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt) }

# create temporary df that includes a log normalized wet weight
temp <- alpha_values_combined %>%
  filter(!is.na(alpha_diff_avg)) %>%
  ungroup() %>%
  mutate(ww_log = log(posttrial_wet_weight_g)) %>%
  select(ww_log,temp_avg)

# plot correlations between log of wet weight and temperature
pairs(temp,lower.panel = panel.smooth, upper.panel = panel.cor,
      gap=0, row1attop=FALSE)
```

## Releveling Lifestage so that Saltwater Subadult is the first reference

-   Here we are releveling the SS stage (saltwater subadult) to be the first reference

-   This does not change the model but changes how we read the output, comparing each stage to the saltwater subadult first.

```{r}
#relevel lifestage so that SS is the first reference
alpha_values_combined$lifestage <- fct_relevel(alpha_values_combined$lifestage, "subadult", "saltwater smolt")

#confused here on why we ran three models but only used model3, I understand that model3 has  more variables so did we deem it best fit/explanatory for the data, if so I think we need to run the model summary statistics here and have an explanation for why we choise model3, or will we leave all three models here (explaining why we will only choose model3 moving forward) and then put model3 in the explanatory section
model1 <- lm(alpha_diff_avg ~ lifestage, alpha_values_combined)
model2 <- lm(alpha_diff_avg ~ lifestage + temp_avg + lifestage:temp_avg, alpha_values_combined)
model3 <- lm(alpha_diff_avg ~ lifestage + temp_avg + t_perc_dw +
               lifestage:t_perc_dw + lifestage:temp_avg, alpha_values_combined)
```

## AIC/BIC Evaluation and Summary Statistics

-   Computing the ACI and BIC to evaluate model complexity and fit

-   The ACI provides an estimate of prediction error, while the BIC also is based on the likelihood function but has a stronger penalty for model complexity

```{r}
#df that shows AIC and BIC results
evaluation <- data.frame(AIC(model1, model2, model3)) %>%
  mutate(BIC = BIC(model1, model2, model3)$BIC) %>%
  arrange(BIC)


plot(model3)
Anova(model3)
summary(model3)
```

## Visualizing the Model

-   This is a visualization of the model, as separated by life stage

```{r}
#visualize the model
library(visreg)
visreg(model3, "temp_avg", by = "lifestage",)
```

-   explain conclusions from plots

## New Data Frame with all Combinations of Lifestage, Temp, and tissue % dry weight

-   Here we are creating a new data frame with all combinations of lifestage, temp, and tissue % dry weight to test model3 predictions

```{r}
# creating new data of all combinations of lifestage, temp, and posttrial_wet_weight
newdata <- expand.grid(lifestage = unique(alpha_values_combined$lifestage),
                       temp_avg = seq(min(alpha_values_combined$temp_avg,na.rm=T),
                                      max(alpha_values_combined$temp_avg,na.rm=T),length.out=100),
                       t_perc_dw = mean(alpha_values_combined$t_perc_dw)
                       )
```

## Testing Model3 on the New Data

```{r}
# how does model3 predict this new data
prdata <- predict(model3,newdata = newdata, se.fit = T)

#creating plot of the new data fit
newdata$fit <- prdata$fit
newdata$lcl <- prdata$fit - prdata$se.fit*2 #lower cl, -fit x2
newdata$ucl <- prdata$fit + prdata$se.fit*2 #upper cl, +fit x2

#plot of this fit with the ucl and lcl
newdata %>%
  ggplot(aes(temp_avg,fit,fill=lifestage)) +
    geom_point(data=alpha_values_combined,aes(temp_avg,alpha_diff_avg,col=lifestage)) +
    geom_line(aes(col=lifestage)) +
    geom_ribbon(aes(ymin = lcl,ymax = ucl),alpha = 0.3)


newdata %>% 
  ggplot(aes(temp_avg, fit, colour = lifestage)) +
    geom_point()
```

-   need to explain here what our conclusions are from these plots

## Plots

-   Here we are creating exploratory plots to determine patterns and relationships and select explanatory outputs

### Plots of alpha by o2 and lifestage

-   Plot alpha vs temp by lifestage

    ```{r}
    #plot alpha vs temp by lifestage
    ggplot(highest_alpha_int, aes(x = temp_c, y = alpha_int_mgo2_kg_h_kPa, color = lifestage)) +
      geom_point(alpha = 0.7)+
      labs(
        title = "Alpha vs Temperature by Lifestage",
        x = "Temperature (°C)",
        y = "Alpha (mgO2kg-1h-1kPa-1)") +
      geom_smooth(method = lm) +
      scale_color_brewer(palette = "Set1")
    ```

-   Plot alpha vs o2 by lifestage

    ```{r}
    #plot alpha vs o2 by lifestage
    ggplot(highest_alpha_int, aes(x = o2_mgl, y = alpha_int_mgo2_kg_h_kPa, color = o2_mgl)) +
      geom_point(alpha = 0.6) +
      scale_color_gradient(low = "blue", high = "red")+
      facet_wrap(~lifestage) +
      labs(
        title = "Alpha vs DO by Lifestage",
        x = "O2 mg/L",
        y = "Alpha (mgO2kg-1h-1kPa-1)") +
      theme_bw()

    ```

-   Plot alpha difference vs temp by lifestage

    ```{r}
    # alpha diff vs temp by lifestage
    ggplot(alpha_values_combined, aes(x = temp_avg, y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

-   plot alpha difference by mass

    ```{r}
    #plot alpha diff by mass
    alpha_values_combined %>% 
      ggplot(aes(x = log(posttrial_wet_weight_g), y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

### Post-mortem plots

-   plot alpha difference by hsi (heposomatic index)

    ```{r}
    #plot alpha diff by hsi
    alpha_values_combined %>% 
      filter(fish_id != "FV077") %>% 
      ggplot(aes(x = hsi, y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

-   plot alpha difference by cf (condition factor)

    ```{r}
    #plot alpha diff by cf
    alpha_values_combined %>% 
      filter(fish_id != "FV091") %>% 
      ggplot(aes(x = cf, y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

-   plot alpha difference by lliver % dry weight

    ```{r}
    #plot alpha diff by l %dw
    ggplot(alpha_values_combined, aes(x = l_perc_dw, y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

-   plot alpha difference by tissue % dry weight

    ```{r}
    #plot alpha diff by tissue %dw
    ggplot(alpha_values_combined, aes(x = t_perc_dw, y = alpha_diff_avg, color = lifestage)) +
      geom_point(alpha = 0.7) +
      scale_color_brewer(palette = "Set2") +
      geom_smooth(method = lm)
    ```

## Summary Statistics

```{r}
#count of each intermittent completed
highest_alpha_int %>% 
  filter(salinity == "Freshwater") %>% 
  count()

highest_alpha_int %>% 
  filter(salinity == "Saltwater") %>% 
  count()

alpha_values_closed %>% 
  filter(salinity == "Freshwater") %>% 
  count()

alpha_values_closed %>% 
  filter(salinity == "Saltwater") %>% 
  count()
```

# Explanatory Data Analysis
